---
title: "Swiss-SEP project"
description: "Aggregating point data to communities"
date: "`r Sys.Date()`"
author:
  - name: Radoslaw Panczak 
    url: https://github.com/RPanczak
    affiliation: ISPM
    affiliation_url: https://www.ispm.unibe.ch/
    orcid_id: 0000-0001-5141-683X
# bibliography: biblio.bib
output:
  distill::distill_article:
    highlight: pygments
    toc: true
    toc_depth: 1
    number_sections: true
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
---

<!-- ------------------------------------------------------------ --> 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      # fig.width=8, fig.height=6, 
                      # out.width="800px", out.height="600px",
                      dpi=300)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
options(scipen=999)
set.seed(12345)
library(pacman) 
p_load(tidyverse, haven, 
       sjmisc, sjPlot,
       sf, tmap)

tmap_mode("view")
```

<!-- ------------------------------------------------------------ --> 

# Raw data 

## Communities

swissBOUNDARIES3D **2018** from [swisstopo](https://shop.swisstopo.admin.ch/en/products/landscape/boundaries3D)

Generalisierte Gemeindegrenzen **2018** from [BfS](https://www.bfs.admin.ch/bfs/de/home/dienstleistungen/geostat/geodaten-bundesstatistik/administrative-grenzen/generalisierte-gemeindegrenzen.html)

```{r include=FALSE}
gem18_st <- readRDS("data/swissBOUNDARIES3D/gem18.Rds") %>%
  st_transform(crs = 2056)

gem18_bfs <- st_read("data-raw/ag-b-00.03-875-gg18/ggg_2018-LV95/shp/g1g18.shp") %>% 
  select(GMDNR)#, GMDNAME)

lake <- st_read("data-raw/ag-b-00.03-875-gg18/ggg_2018-LV95/shp/g1s18.shp")
```

```{r eval=FALSE, include=FALSE}
length(unique(gem18_st$GMDNR))
length(unique(gem18_bfs$GMDNR))

gem18_bfs %>% 
  anti_join(st_drop_geometry(gem18_st)) %>% 
  tm_shape() +
  tm_dots()

tm_shape(gem18_bfs %>% filter(GMDNR == 351)) +
  tm_polygons(col = "green", alpha = 0.5) +
  tm_shape(gem18_st %>% filter(GMDNR == 351)) +
  tm_borders(col = "red", lty = "dotted", alpha = 0.75)
```

Dataset consists of `r scales::number(length(unique(gem18_bfs$GMDNR)), big.mark = ",")` communities.

## Swiss-SEP 

<!-- Using both 1.0 and 2.0 versions -->
Using version 1.0 for now. 

```{r echo=FALSE}
sep1 <- read_rds("data/Swiss-SEP1/ssep_user_geo.Rds") %>% 
  st_transform(crs = 2056) %>% 
  select(v0_buildid, ssep, ssep_d) %>% 
  rename(ssep1 = ssep,
         ssep1_d = ssep_d) %>% 
  mutate(ssep1_d = factor(ssep1_d,
                          levels = 1:10,
                          labels = c("1st - lowest", 
                                     "2", "3", "4", 
                                     "5th decile", 
                                     "6", "7", "8", "9", 
                                     "10th - highest")))

# sep1 %>% 
#   filter(v0_buildid == 982041) %>% 
#   qtm()

sep2 <- read_rds("data/Swiss-SEP2/ssep2_user_geo.Rds") %>% 
  st_transform(crs = 2056) %>% 
  select(gisid, ssep2, ssep2_d) %>% 
  mutate(ssep2_d = factor(ssep2_d,
                          levels = 1:10,
                          labels = c("1st - lowest", 
                                     "2", "3", "4", 
                                     "5th decile", 
                                     "6", "7", "8", "9", 
                                     "10th - highest")))
```

Dataset consists of `r scales::number(nrow(sep1), big.mark = ",")` n'hoods, each assigned to decile of index.

```{r}
st_drop_geometry(sep1) %>% 
  frq(ssep1_d)
```

## STATPOP

```{r include=FALSE}
r18_bu_orig <- read_dta("data-raw/statpop/r18_bu_orig.dta")
# View(r18_bu_orig)

r18_bu <- r18_bu_orig %>% 
  filter(between(r18_gklas, 1110, 1122)) %>% 
  filter(!is.na(r18_gklas)) %>% 
  select(-r18_cant, -r18_cant_sh,
         -r18_ecoord_h, -r18_ncoord_h, 
         -r18_plzz,
         -r18_gklas, -r18_fgkod, -r18_buildper, 
         -r18_nrfloor, -r18_nrdwell, 
         -r18_quarter, -r18_comm, -r18_buildcat) %>% 
  select(-r18_nrpers_main, -r18_nrpers_perm) %>% 
  # filter(!is.na(r18_ncoord)) %>% 
  # filter(!is.na(r18_ecoord)) %>% 
  st_as_sf(coords = c("r18_ecoord", "r18_ncoord"), 
           crs = 2056,
           remove = TRUE)
```

Using data from `STATPOP` **2018**. Excluding `r scales::number(nrow(filter(r18_bu_orig, !between(r18_gklas, 1110, 1122))), big.mark = ",")` nonresidential buildings and `r scales::number(nrow(filter(r18_bu_orig, is.na(r18_gklas))), big.mark = ",")` buildings missing information on classifications. 

Used dataset consists of `r scales::number(nrow(r18_bu), big.mark = ",")` buildings.

<!-- The nearest building with SSEP 1 & 2 were then found and population was aggregated by deciles. -->

The nearest building with SSEP 1 was found; point dataset was then overlaid with PLZ boundaries and summary measures of population based index were created and population level SEP was aggregated by PLZ. 

`r18_nrpers_total` was used for population - slightly different populations would be obtained with different conceptual models of *residential populations* by using either `r18_nrpers_main` or `r18_nrpers_perm` variables


```{r include=FALSE}
rm(r18_bu_orig)
gc()
```

<!-- ------------------------------------------------------------ --> 

# 'nhood based averages 

```{r include=FALSE}
ssep1_gem_18 <- 
  st_join(sep1, gem18_st, join = st_intersects) 

tm_shape(gem18_bfs) +
  tm_polygons() +
  tm_shape(ssep1_gem_18 %>% filter(is.na(GMDNR))) +
  tm_dots()

ssep1_gem_18_neigh <- ssep1_gem_18 %>% 
  st_drop_geometry() %>% 
  relocate(GMDNR, .after = v0_buildid) %>% 
  # covering 4 points outside of the map > lakes & borders
  mutate(GMDNR = ifelse(v0_buildid ==  924650,  158, GMDNR)) %>% 
  mutate(GMDNR = ifelse(v0_buildid ==  982041, 6434, GMDNR)) %>% 
  mutate(GMDNR = ifelse(v0_buildid == 1167211, 5586, GMDNR)) %>% 
  mutate(GMDNR = ifelse(v0_buildid == 1415347, 4806, GMDNR)) %>% 
  group_by(GMDNR) %>% 
  summarise(n = n(),
            ssep1_mean = mean(ssep1),
            ssep1_median = median(ssep1)) %>% 
  ungroup() 

breaks1 <- sep1 %>% 
  st_drop_geometry() %>% 
  group_by(ssep1_d) %>% 
  summarize(min = min(ssep1),
            max = max(ssep1))

# sep1$temp <- cut(sep1$ssep1, breaks = c(breaks1$min, 100),
#                  include.lowest = TRUE,
#                  labels = c("1st - lowest",
#                             "2", "3", "4",
#                             "5th decile",
#                             "6", "7", "8", "9",
#                             "10th - highest"))
# 
# table(sep1$ssep1_d, sep1$temp)
# sep1$temp <- NULL

mapdata_neigh <- inner_join(gem18_bfs, ssep1_gem_18_neigh) %>% 
  filter(!is.na(ssep1_median)) %>% 
  mutate(mean_ssep1_samp_d = ntile(ssep1_mean, 10),
         median_ssep1_samp_d = ntile(ssep1_median, 10)) %>% 
  mutate(mean_ssep1_samp_d = factor(mean_ssep1_samp_d,
                                    levels = 1:10,
                                    labels = c("1st - lowest", 
                                               "2", "3", "4", 
                                               "5th decile", 
                                               "6", "7", "8", "9", 
                                               "10th - highest"))) %>% 
  mutate(median_ssep1_samp_d = factor(median_ssep1_samp_d,
                                      levels = 1:10,
                                      labels = c("1st - lowest", 
                                                 "2", "3", "4", 
                                                 "5th decile", 
                                                 "6", "7", "8", "9", 
                                                 "10th - highest")))

mapdata_neigh$mean_ssep1_full_d <- cut(mapdata_neigh$ssep1_mean, 
                                       breaks = c(breaks1$min, 100),
                                       include.lowest = TRUE,
                                       labels = c("1st - lowest", 
                                                  "2", "3", "4", 
                                                  "5th decile", 
                                                  "6", "7", "8", "9", 
                                                  "10th - highest"))

mapdata_neigh$median_ssep1_full_d <- cut(mapdata_neigh$ssep1_median, 
                                         breaks = c(breaks1$min, 100),
                                         include.lowest = TRUE,
                                         labels = c("1st - lowest", 
                                                    "2", "3", "4", 
                                                    "5th decile", 
                                                    "6", "7", "8", "9", 
                                                    "10th - highest"))

write_rds(mapdata_neigh, "data/mapdata_neigh.rds")
mapdata_neigh %>% 
  st_drop_geometry() %>% 
  write_dta("data/mapdata_neigh.dta")
```

Watch out for small n! Here some examples with communities below 20 n'hoods

```{r}
mapdata_neigh %>% 
  st_drop_geometry() %>% 
  select(GMDNR, n) %>% 
  inner_join(st_drop_geometry(gem18_st) %>% select(GMDNR, NAME)) %>% 
  filter(n < 19) %>% 
  arrange(n)

mapdata_neigh %>% 
  filter(n < 19) %>% 
  qtm(borders = "red", fill = "red")
```

## Final map using median index

Bubbles scaled to number of 'nhoods.

```{r, layout="l-body-outset"}
mapdata_neigh %>% 
  tm_shape() +
  tm_dots(size = "n", col = "ssep1_median", palette = "RdYlGn",
          scale = .5,  shape = 19)
```

## Deciles using sample of PLZs vs 'original' deciles

```{r}
frq(mapdata_neigh$median_ssep1_samp_d)
frq(mapdata_neigh$median_ssep1_full_d)
```

```{r, layout="l-page"}
mapdata_neigh %>% 
  tm_shape() +
  tm_fill(col = c("median_ssep1_samp_d", "median_ssep1_full_d"), 
          border.col = "white",
          palette = "RdYlGn")
```

<!-- ------------------------------------------------------------ --> 

# Population based averages

```{r eval=FALSE, include=FALSE}
# join to sep1
r18_bu_sep <- st_join(r18_bu, sep1, join = st_nearest_feature)

# nearest <- st_nearest_feature(r18_bu, sep1)
# r18_bu_sep$dist1 <- st_distance(r18_bu_sep, sep1[nearest, ], by_element = TRUE)
# rm(nearest)

# join to sep2
r18_bu_sep <- st_join(r18_bu_sep, sep2, join = st_nearest_feature)

# # nearest <- st_nearest_feature(r18_bu_sep, sep2)
# # r18_bu_sep$dist2 <- st_distance(r18_bu_sep, sep2[nearest, ], by_element = TRUE)
# # rm(nearest)

# View(st_drop_geometry(r18_bu_sep))

r18_bu_sep_gem18 <- 
  st_join(r18_bu_sep, gem18_st, join = st_intersects)

# r18_bu_sep_gem18 %>% 
#   filter(is.na(GMDNR))
# 
# tm_shape(gem18_bfs) +
#   tm_polygons() +
#   tm_shape(r18_bu_sep_gem18 %>% filter(is.na(GMDNR)) %>% select(-r18_gdename)) +
#   tm_dots()

r18_bu_sep_gem18 <- r18_bu_sep_gem18 %>% 
  st_drop_geometry() %>% 
  relocate(GMDNR, NAME, .after = r18_egid) %>% 
  # covering 6 points outside of the map > lakes issues
  mutate(GMDNR = ifelse(r18_egid == 899324592, 4806, GMDNR)) %>% 
  mutate(GMDNR = ifelse(r18_egid == 899324589, 4806, GMDNR)) %>% 
  mutate(GMDNR = ifelse(r18_egid == 899324588, 4806, GMDNR)) %>% 
  mutate(GMDNR = ifelse(r18_egid == 898609635,  756, GMDNR)) %>% 
  mutate(GMDNR = ifelse(r18_egid == 898533015, 6417, GMDNR)) %>% 
  mutate(GMDNR = ifelse(r18_egid == 898563492,  942, GMDNR))  

# View(r18_bu_sep_gem18)

write_rds(r18_bu_sep_gem18, "data/statpop/r18_bu_sep_gem18.rds")
```

```{r}
r18_bu_sep_gem18 <- read_rds("data/statpop/r18_bu_sep_gem18.rds") 
```

```{r}
ssep1_gem_18_pop <- r18_bu_sep_gem18 %>% 
  select(-GEM_TEIL, -r18_gdename, -r18_plz4, -ssep1_d, ssep2, ssep2_d, gisid) %>% 
  uncount(r18_nrpers_total) %>% 
  group_by(GMDNR, NAME) %>% 
  summarise(pop = n(),
            ssep1_mean = mean(ssep1),
            ssep1_median = median(ssep1)) %>% 
  ungroup() 

mapdata_pop <- left_join(gem18_bfs, ssep1_gem_18_pop) %>% 
  filter(!is.na(ssep1_median)) %>% 
  mutate(mean_ssep1_samp_d = ntile(ssep1_mean, 10),
         median_ssep1_samp_d = ntile(ssep1_median, 10)) %>% 
  mutate(mean_ssep1_samp_d = factor(mean_ssep1_samp_d,
                                    levels = 1:10,
                                    labels = c("1st - lowest", 
                                               "2", "3", "4", 
                                               "5th decile", 
                                               "6", "7", "8", "9", 
                                               "10th - highest"))) %>% 
  mutate(median_ssep1_samp_d = factor(median_ssep1_samp_d,
                                      levels = 1:10,
                                      labels = c("1st - lowest", 
                                                 "2", "3", "4", 
                                                 "5th decile", 
                                                 "6", "7", "8", "9", 
                                                 "10th - highest")))

mapdata_pop$mean_ssep1_full_d <- cut(mapdata_pop$ssep1_mean, 
                                     breaks = c(breaks1$min, 100),
                                     include.lowest = TRUE,
                                     labels = c("1st - lowest", 
                                                "2", "3", "4", 
                                                "5th decile", 
                                                "6", "7", "8", "9", 
                                                "10th - highest"))

mapdata_pop$median_ssep1_full_d <- cut(mapdata_pop$ssep1_median, 
                                       breaks = c(breaks1$min, 100),
                                       include.lowest = TRUE,
                                       labels = c("1st - lowest", 
                                                  "2", "3", "4", 
                                                  "5th decile", 
                                                  "6", "7", "8", "9", 
                                                  "10th - highest"))

write_rds(mapdata_pop, "data/mapdata_pop.rds")
mapdata_pop %>% 
  st_drop_geometry() %>% 
  write_dta("data/mapdata_pop.dta")
```

## Final map using median index

Bubbles scaled to number of people (2018)

```{r, layout="l-body-outset"}
mapdata_pop %>% 
  tm_shape() +
  tm_dots(size = "pop", col = "ssep1_median", palette = "RdYlGn",
          scale = .5,  shape = 19)
```

## Deciles using sample of communities vs 'original' deciles

```{r}
frq(mapdata_pop$median_ssep1_samp_d)
frq(mapdata_pop$median_ssep1_full_d)
```

```{r, layout="l-page"}
mapdata_pop %>% 
  tm_shape() +
  tm_fill(col = c("median_ssep1_samp_d", "median_ssep1_full_d"), 
          palette = "RdYlGn")
```
